---
title: "p8105_hw3_rl3411"
author: "rl3411"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: github_document
---

```{r, message = F}
library(tidyverse)
library(ggridges)
library(patchwork)
library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "100%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

### Load Instacart data

```{r}
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```

### Understanding the structure of `instacart`

In the `instacart` dataset, there are `r nrow(instacart)` rows and `r ncol(instacart)` columns. Each row corresponds to each item purchased within each order. The variables found in the dataset are:

ID group variables:

1. `order_id`: order identification number
2. `product_id`: product identification number
3. `user_id`: customer identification number
4. `aisle_id`: aisle identification number
5. `department_id`: department identification number

Product group variables:

6. `product_name`: name of the product
7. `aisle`: the aisle name of where the product is
8. `department`: which department the product belongs to

Order group variables:

9. `add_to_cart_order`: order in which each product was added to cart
10. `reordered`: 1 if this product has been ordered by this user in the past, 0 otherwise
11. `order_number`: the order sequence number for this user (1=first, n=nth)
12. `order_dow`: the day of the week on which the order was placed
13. `order_hour_of_day`: the hour of the day on which the order was placed
14. `days_since_prior_order`: days since the last order, capped at 30, NA if `order_number=1`
15. `eval_set`: which evaluation set this order belongs in (train in this case)

Overall, there are `r length(table(pull(instacart, product_id)))` products from the `r length(table(pull(instacart, order_id)))` orders, purchased by `r length(table(pull(instacart, user_id)))`. To get a sense of how this instacart warehouse looks like, we find that there are `r length(table(pull(instacart, department_id)))` departments and `r length(table(pull(instacart, aisle_id)))` aisles.

### Closer look into `instacart`

As aforementioned, there are `r length(table(pull(instacart, aisle_id)))` aisles. Among these aisles, fresh fruits and fresh fruits are two aisles with the most items ordered. This can be seen in the table below. 

```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

Let us first look into the most and least popular aisles. 

```{r}
instacart |> 
  count(aisle) |>
  mutate(aisle = fct_reorder(aisle, n)) |> 
  filter(n > 10000) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Aisles with most orders") + 
  ylab("number of items ordered")
```

Now, let's look into three of the most popular items in aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. In this table, `n` describes the number of times these products have been ordered. Packaged vegetables fruits seem to be the most popular among these three aisles, followed by baking ingredients and dog food care. Among packaged vegetables fruits, organic products seem to be popular, although this is on the assumption that the products provided in that aisle are a good mix between regular and organic items.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Lastly, let us know look at the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week. Generally, we can see that Pink Lady apples are usually ordered around noon or early afternoon, whereas coffee ice cream is ordered later in the afternoon. This pattern holds true on all days of the week, except on Saturdays (or day 5).

```{r, message = F}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```

# Problem 2

### Load BRFSS data

```{r, message = F}
library(p8105.datasets)
data("brfss_smart2010")

brfss_df =
  brfss_smart2010 |> 
  as_tibble() |> 
  janitor::clean_names() |> 
  rename(state = locationabbr,
         county = locationdesc) 
```

### Creating dataset for "Overall Health" topic

```{r}
overall_health_df =
  filter(brfss_df, topic == "Overall Health") |> 
  arrange(factor(response, levels = c("Poor", "Fair", "Good", "Very good", "Excellent")))

overall_health_df
```

### 2002 vs. 2010: States observing at 7 or more locations

```{r, message = F}
locate7_2002 = overall_health_df |> 
  filter(year == "2002") |> 
  count(state) |> 
  arrange(n) |> 
  filter(n >= 7)

locate7_2010 = overall_health_df |> 
  filter(year == "2010") |> 
  count(state) |> 
  arrange(n) |> 
  filter(n >= 7)

compare1 = anti_join(locate7_2002, locate7_2010, by = "state") # all states in 2002 are in 2010
compare2 = anti_join(locate7_2010, locate7_2002, by = "state") # states in 2010 but not in 2002
```

In 2002, `r length(pull(locate7_2002, state))` states observed at 7 or more locations, which were `r pull(locate7_2002, state)`. These states continued observing at 7 or more locations in 2010, and `r length(pull(compare2, state))` other states also started observing at 7 or more locations that year, which are `r pull(compare2, state)`. 

### Spaghetti plot

```{r, message = F}
exc_resp = overall_health_df |> 
  filter(response == "Excellent") |>  
  select(state, year, data_value) |> 
  group_by(state, year) |> 
  summarize(mean = mean(data_value, na.rm=T)) 

exc_resp |> 
  ggplot(aes(x = year, y = mean)) + 
  geom_line(aes(color = (group = state))) + 
  geom_smooth(se = F) + 
  labs(title = "Mean of data values over time (excellent health status group)")
```

From the spaghetti plot above, we see that most of the states have overlapping average values over time for those that responded "excellent" to their general health status. By looking at the thicker smooth blue curve across the center of the plot, there seems to be a slight downward trend from 2002 to 2010 across all states, though the overlaps between each "spaghetti" makes the plot messy and difficult to follow through each state. 

### 2006 vs. 2010: Distribution of data for responses in NY

```{r}
p2c_dataset = overall_health_df |>
  filter(state == "NY",
         year == "2006" | year == "2010") |> 
  select(year, county, response, data_value) 

p2c_dataset |> 
  ggplot(aes(x = factor(response, levels = c("Poor", "Fair", "Good", "Very good", "Excellent")), 
             y = data_value)) +
  xlab("Response") + ylab("Data value") +
  theme(axis.text.x = element_text(size = 8)) +
  geom_boxplot(color = "grey", alpha = 0.5) +
  geom_point(aes(color = county)) + 
  facet_grid(. ~ year) + 
  labs(title = "2006 vs. 2010: Distribution of data for responses in NY")
```

Among locations in NY State in both 2006 and 2010, the distribution of data values seem to increase from "poor" to "very good", then taking a dip in the "excellent" response group. For those who responded "poor", the data value distributions in both years have a small/tight interval and most counties averaged below 5 in this response group. On the other hand, the response groups "good" and "very good" seem to be the highest in both years. In 2006, the distribution of data values seem to be similar between "good" and "very good" but in 2010, the data values for "very good" was apparently higher than those who responded "good".

# Problem 3

### Load, clean and merge dataset

```{r, message = F}
accel_df = 
  read_csv("data/nhanes_accel.csv") |> 
  janitor::clean_names() |> 
  mutate(seqn = as.numeric(seqn)) |> 
  mutate_if(is.double, as.numeric)

demo_df = 
  read_csv("data/nhanes_covar.csv") |> 
  janitor::row_to_names(row_number = 4) |>  # replace variable names
  janitor::clean_names() |>
  mutate(
    sex = case_match( 
      sex,
      "1" ~ "male",
      "2" ~ "female"
    ),
    education = case_match(
      education,
      "1" ~ "less than high school",
      "2" ~ "high school equivalent",
      "3" ~ "more than high school"
      ),
    age = as.numeric(age),
    bmi = as.numeric(bmi),
    seqn = as.numeric(seqn)
    ) |> 
  filter(age >= 21) |> 
  na.omit()

mims_df = 
  inner_join(demo_df, accel_df, by = "seqn") |> 
  arrange(age, bmi)
```

### Comparison of men and women in each education category

```{r}
mims_df |> 
  count(sex, education) |> 
  pivot_wider(
    names_from = education,
    values_from = n 
  ) |>
  relocate("less than high school", .after = "sex") |> 
  knitr::kable(digits = 2)
```


```{r}
mims_df |> 
  ggplot(aes(x = factor(education, 
                        levels = c("less than high school", 
                                   "high school equivalent", 
                                   "more than high school")), 
             y = age, fill = sex)) +
  geom_boxplot() +  
  xlab("Education level") + ylab("Age") +
  labs(title = "Age distribution of men and women in each education category")
```

From the contingency table, we can deduce that most participants have an education level of more than high school. This education group corresponds a lower age distribution in the plot. For those who received an education of less than high school, their age distribution was higher, with an increased median of approximately 20 years as compared to those who had higher level of education. While the numbers between men and women were roughly equal in the "less than high school" and "more than high school" groups, there were 13 more men than women in the group who did high school and equivalent, and men were generally younger than women in this group.

```{r, message = F}
mims_total_act = mims_df |> 
  mutate(total_activity = rowSums(across(min1:min1440))) |> 
  relocate(total_activity, .before = min1) 
  
mims_total_act |> 
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_smooth(aes(fill = sex)) +
  facet_grid(. ~ factor(education, levels = c("less than high school", 
                                              "high school equivalent", 
                                              "more than high school"))) +
  scale_color_viridis_d(option = "H") +
  scale_fill_viridis_d(option = "H") + 
  ylab("Total activity") + xlab("Age") + 
  labs(title = "Total activity levels for men and women of all ages in each education category")
```

Across all three education levels, there is a downward trend of total activity as age increases. This is reasonable as we would expect people to be less active as they grow older. All the curves end between the range of 8000 and 10000 for both men and women at the age of 80, suggesting that the activity levels do not differ between gender for older populations. For those who received the high school equivalent or more than high school, we see that women generally have higher activity levels than men, especially when comparing middle-aged women to men who received "high school equivalent". On the contrary, for those who received "less than high school" education, women are more active than men between 20 to 40 years old, but were less active after the age of 40 as compared to men. Overall, the younger females who received "less than high school" education seem to have the highest level of total activity compared to all other population groups.

```{r, message = F}
timecourse = 
  mims_df |> 
  pivot_longer(
    min1:min1440,
    names_to = "minute_of_the_day",
    values_to = "activity"
  ) |> 
  mutate(minute_of_the_day = str_sub(minute_of_the_day, 4, -1),
         minute_of_the_day = as.numeric(minute_of_the_day)) |> 
  filter(activity < 50)

timecourse |> 
  ggplot(aes(x = minute_of_the_day, y = activity, color = sex)) + 
  geom_point(alpha = 0.1, size = 0.1) + 
  geom_smooth() +
  facet_grid(factor(education, levels = c("less than high school", 
                                          "high school equivalent", 
                                          "more than high school"))~.) +
  scale_x_continuous(breaks = c(0, 250, 500, 750, 1000, 1250, 1500)) +
  xlab("minute of the day") + 
  labs(title = "Activity levels for men and women in each education category across a 24-hour period")
```

Judging by the yellow and purple smooth curves, across all three levels of education, activity levels start to increase at the 375th minute of the day, which corresponds to around 6 o'clock in the morning, then maintain somewhat constant at an average of 12.5 MIMS-units until around 7pm then continues to decrease until midnight. During daytime, women tend to be more active than men, especially those in the "more than high school" education group. As the day passes, we can see that men start to have higher activity levels than women during the night. Lastly, the clusters on the bottom right corner of each plot suggest that people start to rest around 9pm at night in all three groups, assuming that everyone keeps their accelerometers on throughout the 24-hour period. 

